---
title: "Diego-2016-2025-CA-TX-Comparison"
output: html_document
date: "2026-06-01"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r}
#| label: setup

library(tidyverse)
library(janitor)
library(scales)
library(tidycensus)
library(inflateR)
library(jsonlite)

options(scipen = 999)  # Disable scientific notation globally
census_api_key(Sys.getenv("MY_CENSUS_API_KEY"))

agency_categories_json <- read_json("data/agencies_categories.txt")
spending_2020_2024 <- read_csv("data/federal spending by county 2020-2024.csv") |> clean_names()
```

```{r}
#| label: getting-proper-agency-list

agencies_categories <- list()

for (category in names(agency_categories_json)) {
  for (agency in agency_categories_json[[category]]) {
    agencies_categories[[as.character(agency)]] <- category
  }
}
```


Here are the categories for the agencies that have spent money in either Texas or California. I gave Claude the list of agencies and asked it to categorize them by function.

- **Defense and National Security**
  - Department of Defense
  - Department of Homeland Security
  - Nuclear Regulatory Commission
  - Defense Nuclear Facilities Safety Board
  - Selective Service System
- **Social Services & Public Health**
  - Department of Health and Human Services
  - Social Security Administration
  - Department of Housing and Urban Development
  - Department of Veteran Affairs
  - Corporation for National and Community Service
  - Peace Corps
- **Economic & Financial Regulation**
  - Department of the Treasury
  - Securities and Exchange Commission
  - Federal Trade Commission
  - Consumer Financial Protection Bureau
  - Commodity Futures Trading Commission
  - Consumer Product Safety Commission
  - Export-Import Bank of the United States
  - Pension Benefit Guaranty Corporation
  - National Credit Union Administration
  - Small Business Administration
  - Overseas Private Investment Corporation 
  - U.S. International Development Finance Corporation
- **Justice, Law & Civil Rights**
  - Department of Justice
  - Equal Employment Opportunity Commission
  - Federal Election Commission
  - Election Assistance Commission
  - Merit Systems Protection Board
  - Federal Mediation and Conciliation Service
  - National Labor Relations Board
  - Federal Maritime Commission
  - International Trade Commission
- **Science, Technology & Environment**
  - National Aeronautics and Space Administration
  - Department of Energy
  - Environmental Protection Agency
  - National Science Foundation
  - United States Chemical Safety Board
  - Gulf Coast Ecosystem Restoration Council
- **Agriculture, Commerce & Labor**
  - Department of Agriculture
  - Department of Commerce
  - Department of Labor
  - Department of the Interior
  - United States Trade and Development Agency
- **Education, Arts & Culture**
  - Department of Education
  - National Endowment for the Arts
  - National Endowment for the Humanities
  - Institute of Museum and Library Services
  - Smithsonian Institution
  - National Gallery of Art
  - Morris K. Udall and Stewart L. Udall Foundation
- **Infrastructure & Transportation**
  - Department of Transportation
  - National Transportation Safety Board
  - Federal Communications Commission
  - Railroad Retirement Board
- **Foreign Affairs & International Development**
  - Department of State
  - Agency for International Development
  - Millennium Challenge Corporation
  - U.S. Agency for Global Media
- **Government Operations & Administration**
  - General Services Administration
  - Office of Personnel Management
  - Government Accountability Office
  - National Archives and Records Administration
  - Executive Office of the President

# Overall spending comparisons

## Comparing California and Texas in 2016

Adjust for inflation 

```{r}
#| label: adjusting-spending_2020_2024-for-inflation

spending_2020_2024_adjusted <- spending_2020_2024 |> 
  mutate(
    spending_total_adjusted = case_when(
      fyear == 2020 ~ adjust_inflation(spending_total, from_year = 2020, currency = "USD"),
      fyear == 2021 ~ adjust_inflation(spending_total, from_year = 2021, currency = "USD"),
      fyear == 2022 ~ adjust_inflation(spending_total, from_year = 2022, currency = "USD"),
      fyear == 2023 ~ adjust_inflation(spending_total, from_year = 2023, currency = "USD"),
      fyear == 2024 ~ adjust_inflation(spending_total, from_year = 2024, currency = "USD"),
      .default = NA
    )
  ) |>
  select(-c(contracts, grants, direct_payments))

spending_2020_2024_adjusted
```

```{r}
agencies_categories$`Department of Defense`
```


```{r}
#|label: adding-agency-category

spending_2020_2024_adjusted <- spending_2020_2024_adjusted |>
  mutate(
    category = unlist(agencies_categories)[agency]
  )
```


```{r}
#| label: filtering-aggregate-data

spending_2020_tx_ca <- spending_2020_2024_adjusted |> filter(fyear == 2020) |> filter(state_nm == "CALIFORNIA" | state_nm == "TEXAS")
spending_2024_tx_ca <- spending_2020_2024_adjusted |> filter(fyear == 2024) |> filter(state_nm == "CALIFORNIA" | state_nm == "TEXAS")
```

```{r}
#| label: getting-2020-and-2024-population-counts-from-census

# 2020

tx_ca_2020_pop <- get_acs(geography = "state", year = 2020, variables = "B01003_001E", state = "Texas") |> 
  rbind(get_acs(geography = "state", year = 2020, variables = "B01003_001E", state = "California"))

tx_2020_pop <- tx_ca_2020_pop[tx_ca_2020_pop$NAME == "Texas", "estimate"][[1]]
ca_2020_pop <- tx_ca_2020_pop[tx_ca_2020_pop$NAME == "California", "estimate"][[1]]

# 2024

tx_ca_2024_pop <- get_acs(geography = "state", year = 2024, variables = "B01003_001E", state = "Texas") |> 
  rbind(get_acs(geography = "state", year = 2024, variables = "B01003_001E", state = "California"))

tx_2024_pop <- tx_ca_2024_pop[tx_ca_2024_pop$NAME == 'Texas', 'estimate'][[1]]
ca_2024_pop <- tx_ca_2024_pop[tx_ca_2024_pop$NAME == 'California', 'estimate'][[1]]
```

```{r}
#| label: summarizing-2020-federal-spending-in-ca-and-tx

spending_2020_tx <- spending_2020_tx_ca |> 
  filter(state_nm == "TEXAS") 

spending_2020_tx_agency <- spending_2020_tx |> 
  group_by(agency) |>
  summarize(agency_total_tx = sum(spending_total_adjusted), .groups = "drop") |>
  mutate(agency_total_per_capita_tx = round(agency_total_tx / tx_2020_pop, 2)) |>
  arrange(desc(agency_total_per_capita_tx))

spending_2020_ca <- spending_2020_tx_ca |>
  filter(state_nm == "CALIFORNIA") 

spending_2020_ca_agency <- spending_2020_ca |>
  group_by(agency) |>
  summarize(agency_total_ca = sum(spending_total_adjusted), .groups = "drop") |>
  mutate(agency_total_per_capita_ca = round(agency_total_ca / ca_2020_pop, 2)) |> 
  arrange(desc(agency_total_per_capita_ca))

agency_spending_2020_ca_tx <- spending_2020_ca_agency |> 
  full_join(spending_2020_tx_agency, join_by("agency"))

agency_spending_2020_ca_tx

agencies_2020 <- agency_spending_2020_ca_tx |> distinct(agency) |> pull(agency)

all_agency_spending_2020_ca_tx <- agency_spending_2020_ca_tx |>
  summarize(
    ca_all_agency_total_2020 = sum(agency_total_ca, na.rm = T),
    tx_all_agency_total_2020 = sum(agency_total_tx, na.rm = T)
  ) |>
  mutate(
    ca_all_agency_per_capita_2020 = round(ca_all_agency_total_2020 / ca_2020_pop, 2),
    tx_all_agency_per_capita_2020 = round(tx_all_agency_total_2020 / tx_2020_pop, 2)
  )

all_agency_spending_2020_ca_tx
```

```{r}
#| label: summarizing-2024-federal-spending-in-ca-and-tx

spending_2024_tx <- spending_2024_tx_ca |> filter(state_nm == "TEXAS")

spending_2024_tx_agency <- spending_2020_tx |>
  group_by(agency) |>
  summarize(agency_total_tx = sum(spending_total_adjusted), .groups = "drop") |>
  mutate(agency_total_per_capita_tx = round(agency_total_tx / tx_2024_pop, 2)) |>
  arrange(desc(agency_total_per_capita_tx))

spending_2024_ca <- spending_2024_tx_ca |> filter(state_nm == "CALIFORNIA")

spending_2024_ca_agency <- spending_2024_ca |>
  group_by(agency) |>
  summarize(agency_total_ca = sum(spending_total_adjusted), .groups = "drop") |>
  mutate(agency_total_per_capita_ca = round(agency_total_ca / ca_2024_pop, 2)) |>
  arrange(desc(agency_total_per_capita_ca))

agency_spending_2024_ca_tx <- spending_2024_ca_agency |>
  full_join(spending_2024_tx_agency, join_by("agency"))

agency_spending_2024_ca_tx

agencies_2024 <- agency_spending_2024_ca_tx |> distinct(agency) |> pull(agency)

all_agency_spending_2024_ca_tx <- agency_spending_2024_ca_tx |>
  summarize(
    ca_all_agency_total_2024 = sum(agency_total_ca, na.rm = T),
    tx_all_agency_total_2024 = sum(agency_total_tx, na.rm = T)
  ) |>
  mutate(
    ca_all_agency_per_capita_2024 = round(ca_all_agency_total_2024 / ca_2024_pop, 2),
    tx_all_agency_per_capita_2024 = round(tx_all_agency_total_2024 / tx_2024_pop, 2)
  )

all_agency_spending_2024_ca_tx
```

```{r}
#| label: reading-contracts-data

# 2020
tx_contracts_2020 <- read_csv("data/Texas2020Contracts.csv", col_types = cols(.default = "c")) |> 
  clean_names() |> 
  select("contract_award_unique_key","award_id_piid","parent_award_agency_name","current_total_value_of_award","award_base_action_date":"award_latest_action_date_fiscal_year","awarding_agency_name","awarding_sub_agency_name","awarding_office_name","funding_agency_name","funding_sub_agency_name","funding_office_name","foreign_funding","foreign_funding_description","recipient_name","recipient_parent_name","recipient_city_name","recipient_county_name","primary_place_of_performance_city_name","primary_place_of_performance_county_name","award_type","prime_award_base_transaction_description","product_or_service_code_description","domestic_or_foreign_entity","country_of_product_or_service_origin","small_business_competitiveness_demonstration_program","multi_year_contract_code","alaskan_native_corporation_owned_firm":"contracting_officers_determination_of_business_size","emerging_small_business","community_developed_corporation_owned_firm","receives_contracts":"receives_contracts_and_financial_assistance","foreign_owned","small_disadvantaged_business") |> 
  mutate(current_total_value_of_award = as.double(current_total_value_of_award), award_base_action_date = ymd(award_base_action_date), award_latest_action_date = ymd(award_latest_action_date))

ca_contracts_2020 <- read_csv("data/California2020Contracts.csv", col_types = cols(.default = "c")) |>
  clean_names() |> 
  select("contract_award_unique_key","award_id_piid","parent_award_agency_name","current_total_value_of_award","award_base_action_date":"award_latest_action_date_fiscal_year","awarding_agency_name","awarding_sub_agency_name","awarding_office_name","funding_agency_name","funding_sub_agency_name","funding_office_name","foreign_funding","foreign_funding_description","recipient_name","recipient_parent_name","recipient_city_name","recipient_county_name","primary_place_of_performance_city_name","primary_place_of_performance_county_name","award_type","prime_award_base_transaction_description","product_or_service_code_description","domestic_or_foreign_entity","country_of_product_or_service_origin","small_business_competitiveness_demonstration_program","multi_year_contract_code","alaskan_native_corporation_owned_firm":"contracting_officers_determination_of_business_size","emerging_small_business","community_developed_corporation_owned_firm","receives_contracts":"receives_contracts_and_financial_assistance","foreign_owned","small_disadvantaged_business") |>
  mutate(current_total_value_of_award = as.double(current_total_value_of_award), award_base_action_date = ymd(award_base_action_date), award_latest_action_date = ymd(award_latest_action_date))

# 2024
tx_contracts_2024 <- read_csv("data/Texas2024Contracts.csv", col_types = cols(.default = "c")) |> 
  clean_names() |> 
  select("contract_award_unique_key","award_id_piid","parent_award_agency_name","current_total_value_of_award","award_base_action_date":"award_latest_action_date_fiscal_year","awarding_agency_name","awarding_sub_agency_name","awarding_office_name","funding_agency_name","funding_sub_agency_name","funding_office_name","foreign_funding","foreign_funding_description","recipient_name","recipient_parent_name","recipient_city_name","recipient_county_name","primary_place_of_performance_city_name","primary_place_of_performance_county_name","award_type","prime_award_base_transaction_description","product_or_service_code_description","domestic_or_foreign_entity","country_of_product_or_service_origin","small_business_competitiveness_demonstration_program","multi_year_contract_code","alaskan_native_corporation_owned_firm":"contracting_officers_determination_of_business_size","emerging_small_business","community_developed_corporation_owned_firm","receives_contracts":"receives_contracts_and_financial_assistance","foreign_owned","small_disadvantaged_business") |>
  mutate(current_total_value_of_award = as.double(current_total_value_of_award), award_base_action_date = ymd(award_base_action_date), award_latest_action_date = ymd(award_latest_action_date))

ca_contracts_2024 <- read_csv("data/California2024Contracts.csv", col_types = cols(.default = "c")) |>
  clean_names() |> 
  select("contract_award_unique_key","award_id_piid","parent_award_agency_name","current_total_value_of_award","award_base_action_date":"award_latest_action_date_fiscal_year","awarding_agency_name","awarding_sub_agency_name","awarding_office_name","funding_agency_name","funding_sub_agency_name","funding_office_name","foreign_funding","foreign_funding_description","recipient_name","recipient_parent_name","recipient_city_name","recipient_county_name","primary_place_of_performance_city_name","primary_place_of_performance_county_name","award_type","prime_award_base_transaction_description","product_or_service_code_description","domestic_or_foreign_entity","country_of_product_or_service_origin","small_business_competitiveness_demonstration_program","multi_year_contract_code","alaskan_native_corporation_owned_firm":"contracting_officers_determination_of_business_size","emerging_small_business","community_developed_corporation_owned_firm","receives_contracts":"receives_contracts_and_financial_assistance","foreign_owned","small_disadvantaged_business") |>
  mutate(current_total_value_of_award = as.double(current_total_value_of_award), award_base_action_date = ymd(award_base_action_date), award_latest_action_date = ymd(award_latest_action_date))

num_of_tx_2020_contract_awards <- n_distinct(tx_contracts_2020$contract_award_unique_key)
num_of_tx_2024_contract_awards <- n_distinct(tx_contracts_2024$contract_award_unique_key)
num_of_ca_2020_contract_awards <- n_distinct(ca_contracts_2020$contract_award_unique_key)
num_of_ca_2024_contract_awards <- n_distinct(ca_contracts_2024$contract_award_unique_key)
```

```{r}
#| label: more-mutating-contract-data

tx_contracts_2020 <- tx_contracts_2020 |> mutate(minority_owned_business_all = if_else((minority_owned_business == "t" | other_minority_owned_business == "t"), T, F)) 

ca_contracts_2020 <- ca_contracts_2020 |> mutate(minority_owned_business_all = if_else((minority_owned_business == "t" | other_minority_owned_business == "t"), T, F)) 

tx_contracts_2024 <- tx_contracts_2024 |> mutate(minority_owned_business_all = if_else((minority_owned_business == "t" | other_minority_owned_business == "t"), T, F)) 

ca_contracts_2024 <- ca_contracts_2024 |> mutate(minority_owned_business_all = if_else((minority_owned_business == "t" | other_minority_owned_business == "t"), T, F)) 
```


```{r}
#|label: looking-at-2020-minority-business-spending

tx_minority_contracts_2020_agencies <- tx_contracts_2020 |> 
  filter(minority_owned_business_all == T) |>
  group_by(awarding_agency_name) |>
  summarize(
    n_contracts = n(),
    total_agency_award_values = sum(current_total_value_of_award),
    .groups = "drop"
    ) |>
  arrange(desc(total_agency_award_values)) |>
  mutate(
    total_agency_award_per_capita = round(total_agency_award_values / tx_2020_pop, 2),
    category = unlist(agencies_categories)[awarding_agency_name]
  ) 

tx_minority_contracts_2020_overall <- tx_contracts_2020 |>
  group_by(minority_owned_business_all) |>
  summarize(
    n_contracts = n(),
    total_spending = sum(current_total_value_of_award),
    .groups = "drop"
  ) |>
  mutate(
    total_award_spending_per_capita = round(total_spending / tx_2020_pop, 2)
  )

ca_minority_contracts_2020_agencies <- ca_contracts_2020 |>
  filter(minority_owned_business_all == T) |>
  group_by(awarding_agency_name) |>
  summarize(
    n_contracts = n(),
    total_agency_award_values = sum(current_total_value_of_award),
    .groups = "drop"
  ) |>
  arrange(desc(total_agency_award_values)) |>
  mutate(
    total_agency_award_per_capita = round(total_agency_award_values/ ca_2020_pop, 2),
    category = unlist(agencies_categories)[awarding_agency_name]
  ) 

ca_minority_contracts_2020_overall <- ca_contracts_2020 |>
  group_by(minority_owned_business_all) |>
  summarize(
    n_contracts = n(),
    total_spending = sum(current_total_value_of_award),
    .groups = "drop"
  ) |>
  mutate(
    total_award_spending_per_capita = round(total_spending / ca_2020_pop, 2)
  )
```

```{r}
#| label: looking-at-2024-minority-contract-spending

tx_minority_contracts_2024_agencies <- tx_contracts_2024 |> 
  filter(minority_owned_business_all == T) |>
  group_by(awarding_agency_name) |>
  summarize(
    n_contracts = n(),
    total_agency_award_values = sum(current_total_value_of_award),
    .groups = "drop"
    ) |>
  arrange(desc(total_agency_award_values)) |>
  mutate(
    total_agency_award_per_capita = round(total_agency_award_values / tx_2024_pop, 2),
    category = unlist(agencies_categories)[awarding_agency_name]
  ) 

tx_minority_contracts_2024_overall <- tx_contracts_2024 |>
  group_by(minority_owned_business_all) |>
  summarize(
    n_contracts = n(),
    total_spending = sum(current_total_value_of_award),
    .groups = "drop"
  ) |>
  mutate(
    total_award_spending_per_capita = round(total_spending / tx_2024_pop, 2)
  )

ca_minority_contracts_2024_agencies <- ca_contracts_2024 |>
  filter(minority_owned_business_all == T) |>
  group_by(awarding_agency_name) |>
  summarize(
    n_contracts = n(),
    total_agency_award_values = sum(current_total_value_of_award),
    .groups = "drop"
  ) |>
  arrange(desc(total_agency_award_values)) |>
  mutate(
    total_agency_award_per_capita = round(total_agency_award_values/ ca_2024_pop, 2),
    category = unlist(agencies_categories)[awarding_agency_name]
  )

ca_minority_contracts_2024_overall <- ca_contracts_2024 |>
  group_by(minority_owned_business_all) |>
  summarize(
    n_contracts = n(),
    total_spending = sum(current_total_value_of_award),
    .groups = "drop"
  ) |>
  mutate(
    total_award_spending_per_capita = round(total_spending / ca_2024_pop, 2)
  )
```

Comparing spending in 2020

```{r}
#| label: comparing-2020-spending

tx_ca_minority_contracts_overall <- tx_minority_contracts_2020_overall |>
  mutate(state = "Texas", year = 2020) |> 
  rbind(ca_minority_contracts_2020_overall |> mutate(state = "California", year = 2020)) |>
  rbind(rbind(
    tx_minority_contracts_2024_overall |> mutate(state = "Texas", year = 2024), 
    ca_minority_contracts_2024_overall |> mutate(state = "California", year = 2024)
  )) |>
  select(state, year, everything()) |> 
  filter(minority_owned_business_all == T) 

tx_ca_minority_contracts_overall |> write_csv("data-exports/tx-ca-minority-contracts-overall.csv")

tx_ca_minority_contracts_agencies_2020 <- tx_minority_contracts_2020_agencies |> 
  mutate(state = "Texas") |>
  rbind(ca_minority_contracts_2020_agencies |> mutate(state = "California")) |>
  select(state, everything()) |>
  arrange(desc(total_agency_award_per_capita))

tx_ca_minority_contracts_agencies_2020 |> write_csv("data-exports/tx-ca-minority-contracts-agencies-2020-unaggregated.csv")

tx_ca_minority_contracts_agencies_agg_2020 <- tx_ca_minority_contracts_agencies_2020 |>
  group_by(state, category) |>
  summarize(
    n_contracts = sum(n_contracts),
    total_category_awards = sum(total_agency_award_values),
    .groups = "drop"
  ) |>
  mutate(
    total_category_awards_per_capita = case_when(
      state == "Texas" ~ total_category_awards / tx_2020_pop,
      state == "California" ~ total_category_awards / ca_2020_pop
    )
  ) |>
  arrange(desc(total_category_awards_per_capita))

tx_ca_minority_contracts_agencies_agg_2020 |> write_csv("data-exports/tx-ca-minority-contracts-agencies-2020-agg.csv")
  
```

```{r}
#| label: comparing-2024-spending

tx_ca_minority_contracts_agencies_2024 <- tx_minority_contracts_2024_agencies |>
  mutate(state = "Texas") |>
  rbind(ca_minority_contracts_2024_agencies |> mutate(state = "California")) |>
  select(state, everything()) |>
  arrange(desc(total_agency_award_per_capita))

tx_ca_minority_contracts_agencies_2024 |> write_csv("data-exports/tx-ca-minority-contracts-agencies-2024-unaggregated.csv")

tx_ca_minority_contracts_agencies_agg_2024 <- tx_ca_minority_contracts_agencies_2024 |>
  group_by(state, category) |>
  summarize(
    n_contracts = sum(n_contracts),
    total_category_awards = sum(total_agency_award_values),
    .groups = "drop"
  ) |>
  mutate(
    total_category_awards_per_capita = case_when(
      state == "Texas" ~ total_category_awards / tx_2024_pop,
      state == "California" ~ total_category_awards / ca_2024_pop
    )
  ) |>
  arrange(desc(total_category_awards_per_capita))

tx_ca_minority_contracts_agencies_agg_2024 |> write_csv("data-exports/tx-ca-minority-contracts-agencies-2024-agg.csv")
```


```{r}
#| label: finding-a-minority-contractor-in-2024-tx

tx_contracts_2024 |> 
  filter(minority_owned_business_all == T) |>
  group_by(recipient_name, awarding_agency_name) |>
  summarize(
    total_award_sum = sum(current_total_value_of_award),
    .groups = "drop") |>
  mutate(total_award_per_capita = round(total_award_sum / tx_2024_pop, 2)) |> 
  arrange(desc(total_award_per_capita)) |> 
  head(50) 
```

```{r}
tx_contracts_2024 |> filter(recipient_name == "MVM, INC.")
```


